### 2017-08-21 paper/dispRity_paper.tex

The \disp function allows to measure disparity from any matrix (of class \texttt{matrix} or \texttt{data.frame}) where the columns correspond to the different dimensions of the multidimensional space (e.g. the ordination axis or any variable that composes the space) and the rows correspond to the elements present in the multidimensional space (e.g. species, study sites, etc...).

@@

\begin{figure}[!htbp]
\centering
   \includegraphics[width=1\textwidth]{../doc/dispRity_fun.pdf} 
\caption{Illustration of the different metric dimensions levels in the \disp package}
\label{Fig:levels}
\end{figure}

@@

\begin{table}
\center
    \begin{tabular}{r|c|c|c|c}
        name & description & dimension level & definition & source \\
        \hline
        \texttt{ellipse.volume} & The volume of the ellipsoid of the space & 1 & $\frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})$ & \cite{DonohueDim}\\
        \texttt{convhull.surface} & 1 & 1 & 1 & 1\\
        \texttt{convhull.volume} & 1 & 1 & 1 & 1\\
        \texttt{hypervolume} & 0 & 0 & 0 & 1\\
        \texttt{diagonal} & 0 & 0 & 0 & 1\\
        \texttt{ranges} & 0 & 0 & 0 & 1\\
        \texttt{variances} & 0 & 0 & 0 & 1\\
        \texttt{centroids} & 0 & 0 & 0 & 1\\
        \texttt{mode.val} & 0 & 0 & 0 & 1\\
    \end{tabular}
    \caption{Where $k$ is the number of dimensions, $n$ the number of elements, $\lambda_i$ is the eigenvalue of each dimensions, $\sigma^{2}$ is their variance and $Centroid_{k}$ is their mean.}
    \label{Tab:metrics}
\end{table}

% Level | Name | Description | Source | 
% ------|------|---------------------------------------------------|--------|
% 3 | `ellipse.volume`<sup>1</sup> | The volume of the ellipsoid of the space | [@DonohueDim] |
% 3 | `convhull.surface` | The surface of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln` |
% 3 | `convhull.volume` | The volume of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln` |
% 3 | `hypervolume` | The volume of the ordinated space | [`hypervolume`](https://cran.r-project.org/web/packages/hypervolume/index.html)`::hypervolume` |
% 3 | `diagonal` | The longest distance in the ordinated space (like the diagonal in two dimensions) | `dispRity` |
% 2 | `ranges` | The range of each dimension | `dispRity` |
% 2 | `variances` | The variance of each dimension | `dispRity` |
% 2 | `centroids`<sup>2</sup> | The distance between each element and the centroid of the ordinated space | `dispRity` |
% 1 | `mode.val` | The modal value | `dispRity` |

% 1: This function uses an estimation of the eigenvalue that only works for MDS or PCoA ordinations (not PCA).

% 2: Note that by default, the centroid is the centroid of the elements.
% It can, however, be fixed to a different value by using the `centroid` argument `centroids(space, centroid = rep(0, ncol(space)))`, for example the origin of the ordinated space.
%     ellipse.volume = \frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})
%     diagonal = \sqrt{\sum_{i=1}^{k}|max(k_i) - min(k_i)|}
%     ranges = |max(k_i) - min(k_i)|
%     variances = \sigma^{2}{k_i}
%     centroids = \sqrt{\sum_{i=1}^{n}{({k}_{n}-Centroid_{k})^2}}

@@

\texttt{dispRity.through.time(BeckLee\_mat50, )}

This function is a wrapping function that is the equivalent to:

\texttt{dispRity(boot.matrix(time.subsamples(BeckLee\_mat50, )))}

>>>>

The \disp function allows to measure disparity from any matrix (of class \texttt{matrix} or \texttt{data.frame}) where the columns correspond to the different dimensions of the multidimensional space (e.g. the ordination axis or any variable composing the space) and the rows correspond to the elements present in the multidimensional space (e.g. species, individuals, study sites, etc...).

@@

\begin{figure}[!htbp]
\centering
   \includegraphics[width=1\textwidth]{../inst/gitbook/dispRity_fun.pdf} 
\caption{Illustration of the different metric dimensions levels in the \disp package}
\label{Fig:levels}
\end{figure}

@@

\begin{table}
\center
    \begin{tabular}{r|l|c|c|c}
        name & description & dimension level & definition & source \\
        \hline
        \texttt{ellipse.volume}$^1$ & The volume of the ellipsoid around the elements & 1 & $\frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})$ & \cite{DonohueDim}\\
        \texttt{convhull.surface} & The surface of the convex hull formed by the elements & 1 & NA & \texttt{geometry::convhulln} \\
        \texttt{convhull.volume} & The volume of the convex hull formed by the elements & 1 & NA & \texttt{geometry::convhulln} \\
        \texttt{hypervolume} & The volume of the multidimensional space space  & 1 & NA & \texttt{hypervolume::hypervolume} \\
        \texttt{diagonal} & The longest Eucliden distance between any pair of elements & 1 & $\sqrt{\sum_{i=1}^{k}|max(k_i) - min(k_i)|}$ & \texttt{dispRity::diagonal} \\
        \texttt{mode.val} & The modal value of a vector or a matrix & 1 & NA & \texttt{dispRity::mode.val}\\
        \texttt{ranges} & The absolute ranges of each dimension & 2 & $|max(k_i) - min(k_i)|$ & \texttt{dispRity::ranges} \\
        \texttt{variances} & The variance of each dimension & 2 & $\sigma^{2}{k_i}$ & \texttt{dispRity::variances} \\
        \texttt{centroids} & The distance between each element and a fixed point$^2$ of the space & 2 & $\sqrt{\sum_{i=1}^{n}{({k}_{n}-Centroid_{k})^2}}$ & \texttt{dispRity::centroids} \\
    \end{tabular}
    \caption{Where $k$ is the number of dimensions, $n$ the number of elements, $\lambda_i$ is the eigenvalue of each dimensions, $\sigma^{2}$ is their variance and $Centroid_{k}$ is their mean. $^1$ this function uses a fast estimation of the eigenvalue that only works in an ordinated space based on MDS or PCO/PCoA (\textit{not} PCA); $^2$ by default that point is the centroid of the elements.}
    \label{Tab:metrics}
\end{table}

@@

As an example, one can define their multidimensional space as being the morphospace of an ordination of the distance between discrete morphological characters for 50 mammals species \citep[from][]{beckancient2014}.
Then, one can define disparity as the sum of the variances on each dimensions \citep{Wills1994} that will represent an aspect of the the volume of the morphospace (i.e. a large sum of variances approximates a large morphospace).
One can then be interested in how this approximation of the volume of the morphospace changes throughout different time periods (say from the end of the Cretaceous to the end of the Palaeogene).

\texttt{disparity <- dispRity.through.time(data = BeckLee\_mat50, time = c(90, 70, 50, 30), metric = c(sum, variances))}
+ method = discrete
+ tree = tree

\noindent Note that this function is a wrapping function that is the equivalent to:

\texttt{disparity <- dispRity(boot.matrix(time.subsamples(data = BeckLee\_mat50, time = c(90, 70, 50, 30))), metric = c(sum, variances))}

\noindent Which allows a finer tuning of each optional argument in each of the three functions \texttt{time.subsamples}, \texttt{boot.matrix} and \disp.
The three arguments here are pretty self explanatory: \texttt{data = BeckLee\_mat50} is our definition of the multidimensional space, \texttt{time = c(90, 70, 50, 30)} indicates the boundaries of the different time bins and \texttt{metric = c(sum, variances)} is our definition of disparity.

This function returns a \disp object that does not directly gives us the results (the disparity in each time bin) but rather summarises the content of the \disp object.
To visualise the disparity values, one can use the \texttt{summary} or/and \texttt{plot} options:

\texttt{summary(disparity)}

\texttt{plot(disparity)}
